Deep Reinforcement Learning for Conservation Decisions

### 2021-05-05 manuscript/manuscript.Rmd
Passed over intro

<!-- Environmental problems, ML -->
Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
But will ML approaches really help us tackle a changing planet?
Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* (e.g. image recognition) and *unsupervised learning* (including clustering and prediction), but have so far largely overlooked the third and possibly most promising approach in the ML triad: Reinforcement Learning (RL).
Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:

@@

Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games.
RL agents have recently achieved superhuman performance on classic Atari games [@DQN], while Deepmind’s RL-trained AlphaZero has defeated the world's grandmasters in Go, Chess, and Shogi [@alphazero].
Complex environmental problems share many similarities to these games: the need to plan many moves ahead, accept short-term costs to achieve long-term benefit, to account for uncertainty and respond with contingency to the unexpected.
Of course real ecosystems are also far more complex than a game of chess or MoonLander -- but can the same be said of our efforts at finding the best strategy?
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management. 
Even simpler models are frequently employed throughout ecology to guide understanding and response.
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of models, the so called "curse of dimensionality" makes problems with too many possible states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
Neural-network-based ML techniques have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
We believe that the combination of such approaches from RL trained using sufficient simulations of suitable ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.

@@

Any application of RL can be divided into two components: an *environment* and an *agent*.
An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
An *agent* is typically a computational neural network based on an appropriate RL algorithm that may have been 'trained' on one or more environments.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
In principle, there are virtually no restrictions on what comprises a state: it may be continuous or discrete, completely or only partially observed, one or many dimensions.
For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
The reward will in general depend on both the current state and choice of actions.
An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.

@@

The training an agent on an environment involves allowing the agent to experience and respond to repeated simulations from the environment, thousands or even millions of times. 
Over the course of training, weights on the neural network are updated in response to the reward given by the environment, much like reinforcement learning is believed to work in living creatures [@cite].
A neural network can be thought of as a generic function approximator, and these weights as parameters of that function that are estimated from 'data' generated by training.
A typical neural network in RL may have thousands or hundreds of thousands of parameters, giving it the flexibility to learn very complex behaviors.
The overall architecture of the neural network, as well as other parameters such as those governing how an agent learns, how an agent proposes actions, and how the agent discounts future rewards, and so forth are controlled by *hyperparameters*.
To initialize a new, untrained agent, we must specify a handful or so of these hyperparameters, which are held fixed over the course of training.
The particular selection of hyperparameters required depends on which *RL algorithm* our agent will be based.
The RL algorithm specifies precisely how the agent will learn from its environment:
Is the algorithm model-based or model-free? 
Stochastic or deterministic? 
"Off-policy" or "on-policy"? 
Does it use a random noise model to propose actions?
Does it permit a continuous action space or require discrete actions?
The design of RL algorithms has been and remains an area of intense research and rapid development [Table 1].

The process of selecting hyperparameters for a given RL algorithm to maximize it's performance on a given environment is referred to as hyperparameter *tuning*.
The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
Because training even on a fixed set of hyperparameters can be computationally intensive, tuning must be targeted appropriately.
Considerations in environmental design, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.

@@

# RL essentials

@@

Meanwhile, the state-action-value or quality function, finds the expected return given a state and action, $$
  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
$$

@@

<!-- Can we add citations to each please -->

```{=latex}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Algorithm & Description & Model & Policy & Action Space & State Space \\
\hline
MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
\hline
MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
\hline
A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
\hline
A3C  & Asynchronous A2C & Model-free & On-policy & &\\
\hline
TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
\hline
PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
\hline
DQN & Deep Q Networks & Model-free & Off-policy& &\\
\hline
DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
\hline
TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
\hline
SAC & Soft Actor Critic & Model-free & Off-policy & &\\
\hline
IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
\hline
APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
\hline
\end{tabular}
\end{center}
```
# Examples

>>>>

<!-- Environmental problems, ML -->
Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
But will ML approaches really help us tackle a changing planet?
Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:

@@

Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
RL agents develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is a common practice.
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management. 
Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.

@@

Any application of RL can be divided into two components: an *environment* and an *agent*.
An *environment* is typically a computer simulation, though it is certainly possible to learn from real world environments, such as teaching physical robots to walk and navigate obstacles.
An *agent* is typically a computer algorithm which takes sequential observations from the environment and returns proposed actions. RL agents typically use one or more neural networks to represent the relationship between observed states and suitable actions, and 'train' these networks through repeated interactions with the environment to maximize the culumulative reward.
<!-- Blah blah skip this?-->
It is important to recognize that these definitions are significantly more general than requirements typically imposed by classical methods.  
For instance, there is no a priori assumption that the environment is *fully observed*, or that the environment is *Markovian* (future states depend only on the current state), commonly assumed by classical methods [e.g. @Marescot2013]. 
Just as important is the high degree of interoperability provided by the RL framework.
As we shall see, an RL algorithm developed to play Atari games can instead be trained on an environment that simulates fish population dynamics without modification.
(Note that this does not refer to the *trained* agent).
This not guarantee that such an algorithm will be successful (even after extensive tuning and training in the fish population dynamics environment), but does open the door to more rapid exchange between fields.
In particular, the adoption of standard interfaces for RL environments [@brockman2016] as well as the implementation of cutting edge algorithms in flexible open source frameworks [@sb3; @tf-agents] means that a researcher can test a range of RL algorithms against a single environment, or a given algorithm across a range of environments with equal ease.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
In contrast to classical approaches, there are few restrictions on what comprises a state or action, at least in principle.
The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
The reward will in general depend on both the current state and choice of actions.
An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{fig1.pdf}
\caption{Reinforcement Learning focuses on the repeated feedback of agent interacting with an \emph{environment}. The agent performs an action which influences the environment's future \emph{state}.  The agent receives an observation of the new state and reward from the environment.}
\end{figure}

@@

The training an agent on an environment involves allowing the agent to experience and respond to repeated simulations from the environment, thousands or even millions of times. 
Over the course of training, weights on the neural network are updated in response to the reward given by the environment, much like reinforcement learning is believed to work in living creatures [@cite].
A neural network can be thought of as a generic function approximator, and these weights as parameters of that function that are estimated from 'data' generated by training.
A typical neural network in RL may have thousands or hundreds of thousands of parameters, giving it the flexibility to learn very complex behaviors.
The overall architecture of the neural network, as well as other parameters such as those governing how an agent learns, how an agent proposes actions, and how the agent discounts future rewards, and so forth are controlled by *hyperparameters*.
To initialize a new, untrained agent, we must specify a handful or so of these hyperparameters, which are held fixed over the course of training.
The particular selection of hyperparameters required depends on which *RL algorithm* our agent will be based.
The RL algorithm specifies precisely how the agent will learn from its environment:
Is the algorithm model-based or model-free? 
Stochastic or deterministic? 
"Off-policy" or "on-policy"? 
Does it use a random noise model to propose actions?
Does it permit a continuous action space or require discrete actions?
The design of RL algorithms has been and remains an area of intense research and rapid development [Table 2].
Open source implementations by professional software development teams often follow swiftly after new algorithms are published [@Ssb3; @tf-agents].
In the following section, we summarize some of the key concepts that make up an RL algorithm, and highlight relevant distinctions between important algorithms today.

<!-- Something about given RL algorithm being able to solve lots of different environments with no pre-specified knowledge. cite examples @AlphaZero and @DQN; but still depends on hyperparameters -->
The process of selecting hyperparameters for a given RL algorithm to maximize it's performance on a given environment is referred to as hyperparameter *tuning*.
The process of tuning can be automated through standard multi-dimensional optimization approaches: proposing new combinations of hyperparameters, training an agent, and examining the resulting performance.
Because training even on a fixed set of hyperparameters can be computationally intensive, tuning must be targeted appropriately.
Considerations in how the environment simulation is implemented, such as the normalization of state and/or action spaces, can significantly improve performance of many RL algorithms on a diverse range of problems without requiring extensive tuning.

@@

# RL Algorithms

I think we can lean more heavily both in references to and paralleling the presentation of some of the existing background material.

@@

Meanwhile, the state-action-value or quality function, finds the expected return given a state and action,

$$
  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
$$

@@

<!-- Can we add citations to each please -->
<!--

Should we mention:

- Temporal Differencing, RIENFORCE
- SARSA, 
- Double DQN?  
Can we add HER (since delayed reward case may be particularly interesting)
Should we do anything to better reflect the 'family tree' here? (e.g. DQN -> DDPG -> TD3)

-->
```{=latex}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Algorithm & Description & Model & Policy & Action Space & State Space \\
\hline
MBPO & Model-based Policy Optimization & Model-based & On-policy& &\\
\hline
MCTS & Monte Carlo Tree Search & Model-based & Either & & \\
\hline
A2C  & Advantage Actor Critic & Model-free & On-policy& &\\
\hline
A3C  & Asynchronous A2C & Model-free & On-policy & &\\
\hline
TRPO  & Trust Region Policy Optimization & Model-free & On-policy & &\\
\hline
PPO  & Proximal Policy Optimization & Model-free & On-policy & &\\
\hline
DQN & Deep Q Networks & Model-free & Off-policy& &\\
\hline
DDPG & Deep Deterministic Policy Gradient & Model-free & Off-policy & &\\
\hline
TD3 & Twin Delayed DDPG & Model-free & Off-policy & &\\
\hline
SAC & Soft Actor Critic & Model-free & Off-policy & &\\
\hline
IMPALA & Importance Weighted Actor Learner & Model-free & Off-policy & &\\
\hline
APE-X & Distributed Prioritized Experience Replay & Model-free & Off-policy & & \\
\hline
\end{tabular}
\end{center}
```
# Examples
